---
title: "TO 414 Team Project 1"
author: "Simon Verghese"
date: "2/19/2021"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library("ggplot2")
library("leaflet")
```

```{r}
#read in data
hotels = read.csv('hotel_bookings.csv', stringsAsFactors = TRUE)
```

Questions to Answer:

Intro:
  Who is the report going to?
  What is the overall goal of the report?
  
Content:
  How should we spend our marketing budget?
    What customer segments should we invest more into marketing to?
    What are the demographics of our most popular customer?
    What countries bring in the most revenue?
    What distribution channel produces the most revenue?
    Should we prioritize company bookings over individual bookings?
  How should we manage bookings and staff?
    What are our busiest times of the year?
    How much does demand change from weekdays to weekends?
    Should we overbook the hotel? If so, by how much?
  What areas of our business need the most improvement?
    How should we address the high cancellation rate?
    How can we get more repeat customers?
    
Conclusion:
  What did we find out?
  What is our final recommendations?





##What is the cancellation rate for the hotels?

```{r}
cancellation_rate = round(sum(hotels$is_canceled)/nrow(hotels), 2)
cancellation_rate
```
The cancellation rate across all the hotels is 37%.


##What month is the most popular?

```{r}
month_arrivals = sort(table(hotels$arrival_date_month))
month_arrivals = data.frame(month_arrivals)
ggplot(data = month_arrivals, aes(x = Var1,y = Freq)) + geom_bar(stat = "identity")
``` 

From the graph we can see that May, July, and August were the most popular months for both types of hotels.

##How many companies came to the hotels?

```{r}
nlevels(hotels$company)
```
353 companies came to the hotels.

##How many total bookings were from agencies?

```{r}
nrow(hotels) - summary(hotels$agent)[2]
```
103,050 bookings were made through agenceis.

##Which countries had the most reservations?

```{r}
country_arrivals = sort(table(hotels$country), decreasing = TRUE)
country_arrivals = data.frame(country_arrivals)[1:10,]
country_arrivals
ggplot(data = country_arrivals, aes(x = Var1,y = Freq)) + geom_bar(stat = "identity")
```

People from PRT, GBR, and FRA had the most bookings.

